---
title: "NeuralNetwork_Novikov"
author: "Vitalii Novikov"
format: pdf
---



# Libraries

```{r}
#| code-summary: Libraries
#| code-fold: true
library <- function(...) {suppressPackageStartupMessages(base::library(...))}
if (!require(caret)) install.packages("caret"); library(caret)
if (!require(neuralnet)) install.packages("neuralnet"); library(neuralnet)
if (!require(mlbench)) install.packages("mlbench"); library(mlbench)
if (!require(dplyr)) install.packages("dplyr"); library(dplyr)
```

# Starting point

We would like to use kNN method for recognizing the type of iris (class label *versicolor* or *virginica*). The predictors include:

- Sepal.Length
- Sepal.Width
- Petal.Length
-Petal.Width

The objective of the analysis is to find a model that can be used to predict the type of the flower based on features. Also we will use extensive search approach to identify a good k for the task.


# Data management

```{r}
data("BreastCancer", package = "mlbench")
BreastCancer |> head()
```


